FastSRM: A fast, memory efficient and identifiable implementation of the shared response model

### 2021-11-25 paper/main.tex
update readme
update ppaer

In this work, we introduce FastSRM, an identifiable model that uses an optimal
atlas-based decomposition to speed up the computations with provably no loss of
performance.
%
FastSRM gains identifiability by imposing that the covariance of the shared
components is diagonal (see ~\cite{richard2020modeling} Appendix D for a proof).
%
The gains in speed and memory usage are based on the intuition that a compressed
representation of the input should be good enough to find a suitable estimate of
the shared components as they live in reduced space.
%
It turns out that there
exists an optimal atlas that reduces the data in a such a way that the obtained
shared components (and other parameters) are the same as if they were computed
on full data.
%
We show that FastSRM brings several practical benefits: on real data, it produces more stable
estimates of the shared components than its non-identifiable counter-part and 
 it is much faster and more memory
efficient than other implementations that do not make use of the optimal atlas.
% on synthetic as well as on real data,

@@

A complexity analysis shows that finding the optimal atlas becomes the limiting
step of the pipeline.
%
Even with fast implementations, the subject specific PCA
is costly.
%
However FastSRM only works on $\zb_i$ so we do not need to know the value of
$U_{\xb_i}$.
%
In practice, we observe data $X_i \in \RR^{v \times n}$ and we want to get $Z_i
\in \RR^{n \times n}$ such that $X_i = U_{\xb_i} Z_i$.
%
This can be done by performing an
SVD of $X_i^{\top} X_i$ yielding $X_i^{\top}X_i= V_i D_i V_i^{\top}$ and setting
$Z_i = D_i^{\frac12} V_i^{\top}$.
%
Although computing the product $X_i^{\top} X_i$ has time complexity
$\bigO(vt^2)$ (the same as a PCA) the constant in the $\bigO$ is exactly one so
this operation costs a lot less than the PCA
on full data.
%
When estimates of the mixing matrices are needed, they can be obtained by
applying equation~\eqref{eq:fastdetsrm:Aiupdate} in the deterministic SRM case and
equation~\eqref{eq:fastprobsrm:Aiupdate} in the probabilistic SRM case which only costs
$\bigO(mvp^2)$.
%
In practice the cost of the matrix products $X_i^{\top} X_i$ is often still the
limiting step of the pipeline (this depends on the number of iterations) but as
we show next, it is much more efficient than performing SRM on
the full data.
%
Note that if memory allows it, these matrix products can be
computed in parallel.
%

@@

We empirically see that the optimal approach is equivalent to using no atlas in
terms of performance.
%
This is predicted by the theory in section~\ref{sec:optimal_atlas} where we demonstrate that these two algorithms yield exactly the same output
from the same input.
%
% %
% Among approximate methods, the projection
% method yields the best results but are quite far away from optimal results.
%
% %
In general probabilistic methods give much better results than their deterministic
counterpart.
%
This shows the superiority of likelihood based methods.
%
In terms of fitting time, FastSRM is about a thousand time faster than SRM after
100 iterations.
%
When no atlas is used, the number
of iterations has a very strong impact on performance while it has a small impact
when the optimal atlas is used.
%
Lastly, looking at the convergence curves, we see that even after $100$ iterations, algorithms did
not fully converge.
%
This means that in practice a larger number of
iterations is needed, which would yield an even higher difference in fitting time
between methods using no atlas and methods using the optimal atlas.
%

@@

% 
We use 5-Fold cross-validation across runs: the training set contains 80\% of
the runs and the test set 20\%, and repeat the experiment using all possible
choices for left-out subjects.
%
% 
The mean accuracy is reported in Figure~\ref{fig:srm:timesegment} (top).
%
When the optimal atlas is used, the accuracy is the same as when no
atlas is used but the fitting time is reduced by a factor 10 to 100
and so is the memory usage (see Figure~\ref{fig:srm:timesegment},
bottom).
%

@@

The package comes with a documentation and working examples (

@@

%
FastSRM is an efficient implementation of SRM
that uses optimal atlases to speed up computations and reduce memory requirements
with provably no loss of performance.

We show on synthetic and real data that FastSRM is much faster and more memory
efficient that implementations not using the optimal atlas.
Furthermore, FastSRM is identifiable. On real data, we show that compared to
Brainiak's implementation (which is not identifiable), FastSRM provides more
stable estimates of the shared components.

%
FastSRM inherits from all the weaknesses of SRM including the fact that
mixing matrices are assumed to be orthogonal which is rather unrealistic. It
remains to be seen whether dimension reduction and learning of the shared
components could be done jointly and efficiently without assuming orthogonal
mixing matrices.
%
Our optimal atlas apprach shows that SRM-like methods do not need the
full data to provide accurate estimates of shared components. We believe such
insights may guide the design of future methods.
\bt{Add a paragraph on implementation, sofwtare quality, documentation, examples, availability...}
\bt{Discuss the use for other brain imaging modelities (MEEG) if relevant ?}

>>>>

In this work, we introduce FastSRM, an identifiable model that uses an optimal
atlas-based decomposition to speed up the computations with provably no loss of
performance.
%
FastSRM gains identifiability by imposing that the covariance of the shared
components is diagonal (see ~\cite{richard2020modeling} Appendix D for a proof).
%
The gains in speed and memory usage are based on the intuition that a compressed
representation of the input should be good enough to find a suitable estimate of
the shared components as they live in reduced space.
%
It turns out that there
exists an optimal atlas that reduces the data in a such a way that the obtained
shared components (and other parameters) are the same as if they were computed
on full data.
%
We show that FastSRM brings several practical benefits: on real data, it produces more stable
estimates of the shared components than its non-identifiable counter-part and 
 it is much faster and more memory
efficient than other implementations that do not make use of an optimal atlas.
% on synthetic as well as on real data,

@@

A complexity analysis shows that finding an optimal atlas becomes the limiting
step of the pipeline.
%
Even with fast implementations, the subject specific PCA
is costly.
%
However FastSRM only works on $\zb_i$ so we do not need to know the value of
$U_{\xb_i}$.
%
In practice, we observe data $X_i \in \RR^{v \times n}$ and we want to get $Z_i
\in \RR^{n \times n}$ such that $X_i = U_{\xb_i} Z_i$.
%
This can be done by performing an
SVD of $X_i^{\top} X_i$ yielding $X_i^{\top}X_i= V_i D_i V_i^{\top}$ and setting
$Z_i = D_i^{\frac12} V_i^{\top}$.
%
Although computing the product $X_i^{\top} X_i$ has time complexity
$\bigO(vt^2)$ (the same as a PCA) the constant in the $\bigO$ is exactly one so
this operation costs a lot less than the PCA
on full data.
%
When estimates of the mixing matrices are needed, they can be obtained by
applying equation~\eqref{eq:fastdetsrm:Aiupdate} in the deterministic SRM case and
equation~\eqref{eq:fastprobsrm:Aiupdate} in the probabilistic SRM case which only costs
$\bigO(mvp^2)$.
%
In practice the cost of the matrix products $X_i^{\top} X_i$ is often still the
limiting step of the pipeline (this depends on the number of iterations) but as
we show next, it is much more efficient than performing SRM on
the full data.
%
Note that if memory allows it, these matrix products can be
computed in parallel.
%

@@

We empirically see that the optimal approach is equivalent to using no atlas in
terms of performance.
%
This is predicted by the theory in section~\ref{sec:optimal_atlas} where we demonstrate that these two algorithms yield exactly the same output
from the same input.
%
% %
% Among approximate methods, the projection
% method yields the best results but are quite far away from optimal results.
%
% %
In general probabilistic methods give much better results than their deterministic
counterpart.
%
This shows the superiority of likelihood based methods.
%
In terms of fitting time, FastSRM is about a thousand time faster than SRM after
100 iterations.
%
When no atlas is used, the number
of iterations has a very strong impact on performance while it has a small impact
when an optimal atlas is used.
%
Lastly, looking at the convergence curves, we see that even after $100$ iterations, algorithms did
not fully converge.
%
This means that in practice a larger number of
iterations is needed, which would yield an even higher difference in fitting time
between methods using no atlas and methods using an optimal atlas.
%

@@

% 
We use 5-Fold cross-validation across runs: the training set contains 80\% of
the runs and the test set 20\%, and repeat the experiment using all possible
choices for left-out subjects.
%
% 
The mean accuracy is reported in Figure~\ref{fig:srm:timesegment} (top).
%
When an optimal atlas is used, the accuracy is the same as when no
atlas is used but the fitting time is reduced by a factor 10 to 100
and so is the memory usage (see Figure~\ref{fig:srm:timesegment},
bottom).
%

@@

The package comes with a documentation and a tutorial that reproduces, at a
smaller scale, the experiment available in Figure~\ref{fig:srm:synthetic_gradient}.
These are found at \url{https://hugorichard.github.io/FastSRM/index.html}.
The instructions to fully reproduce the experiments presented in the papers are
available in the README:
\url{https://github.com/hugorichard/FastSRM/blob/master/README.md}.

A set of tests reduces the risk of introducing a bug. These tests are run any
time a pull request or a merged is performed on the main codebase.

@@

%
FastSRM is an efficient implementation of SRM
that uses optimal atlases to speed up computations and reduce memory requirements
with provably no loss of performance. 
Note that the new memory requirements do not depend on the number of subjects.
This is really interesting as it would make it possible to apply FastSRM on
datasets with a very large number of subjects that do not hold in memory.

We show on synthetic and real data that FastSRM is much faster and more memory
efficient that implementations not using an optimal atlas.
Furthermore, FastSRM is identifiable. On real data, we show that compared to
Brainiak's implementation (which is not identifiable), FastSRM provides more
stable estimates of the shared components.

%
FastSRM inherits from all the weaknesses of SRM including the fact that
mixing matrices are assumed to be orthogonal which is rather unrealistic. It
remains to be seen whether dimension reduction and learning of the shared
components could be done jointly and efficiently without assuming orthogonal
mixing matrices.
%
Our optimal atlas approach shows that SRM-like methods do not need the
full data to provide accurate estimates of shared components. We believe such
insights may guide the design of future methods.
%

Lastly, FastSRM is useful when the number of features is much larger than the
number of samples. This setting is common in fMRI but less common in MEG/EEG where
the number of samples is usually much larger than the number of features. Even
in this case, it is possible to considerably speed up SRM. Indeed, SRM only
depends on second order statistics.
These statistics can be pre-computed making the rest of the
pipeline insensitive to the number of samples in the dataset. An implementation
of this simple idea would make SRM easily applicable to other modalities.
We live it to future work.
% \bt{Add a paragraph on implementation, sofwtare quality, documentation, examples, availability...}
% \bt{Discuss the use for other brain imaging modelities (MEEG) if relevant ?}
